---
title: "A quick tour of ggplot2"
author: "STA313 Teaching Team"
date: "September, 2026"
output: html_document
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install and load packages required for the tutorial
if (!require(palmerpenguins)) {
  install.packages("palmerpenguins") # install the package if missing
}
if (!require(mosaicData)) {
  install.packages("mosaicData") # install the package if missing
}
# library(ggplot2)  # plotting
# library(dplyr)    # data manipulation
library(tidyverse)  # load them together (and more)
library(palmerpenguins) # palmer penguins dataset
library(mosaicData)     # SAT dataset (and more)
```

I am loading all required libraries in the `setup` code chunk above to 

1. ensure all required libraries are available throughout the document; and
2. avoid loading the packages multiple times unnecessarily.

Alternatively, you can load a library in a separate code chunk just before you 
start using it. 

_Note: Loading `tidyverse` also loads `ggplot2`. We will also use functions
from the `tidyverse` family which are not included in `ggplot2`. Alternatively, 
you can individually load `ggplot2` and `dplyr` which contains the other 
required functions._


## Palmer penguins

Measurements for penguins species, island in Palmer Archipelago, size 
(flipper length, body mass, bill dimensions), and sex.

_Hint: To run one chunk at a time, press "Cmd + Shift + Return" or 
"Ctrl + Shift + Enter" with the typing cursor inside the chunk. To run one line 
at a time, press "Cmd + Return" or "Ctrl + Enter" with the typing cursor on 
the line of the code._

```{r}
glimpse(penguins)
```

```{r}
# create the Palmer Penguins plot here
```

## Exercise 1

First, have a look at the data set. Alternatively, type `?mpg` in the console to 
load the documentation page for more details after loading `tidyverse`.

```{r}
glimpse(mpg) # mpg and glimpse are available through `tidyverse`
```


```{r}
# complete Exercise 1 here
```


## Exercise 2

```{r}
# complete Exercise 2 here
```

## Exercise 3

```{r}
# complete Exercise 3 here
```


## Exercise 4

```{r eval=FALSE}
# included for demonstration purposes but not executed
# the data is loaded at the top of the document already
library(mosaicData)
```

First, have a look at the data set. Alternatively, type `?SAT` in the console to 
load the documentation page for more details after loading `mosaicData`.

```{r}
glimpse(SAT) # SAT is available through `mosaicData`
```

```{r}
# complete Exercise 4 here
```

You can save the last plot generated to a file using `ggsave()`.

```{r eval=FALSE}
ggsave("sat.png")
```

## In-tutorial Exercise

```{r eval=FALSE}
mutate(mpg, class = forcats::fct_infreq(class)) |>
  ggplot() +
  geom_bar(
           position = "fill") +    # try using "dodge" and "stack" (default)
  scale_fill_viridis_d()           # see ?scale_fill_viridis_d
```

