---
title: "Tutorial 4"
author: "STA313"
date: "2024-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
packages <- c("patchwork", "ggrepel", "ggtext")
# install missing packages
for (p in packages) {
  if (!require(p, character.only = TRUE)) {
    install.packages(p)
    library(p, character.only = TRUE)
  }
}
```

# Code on slide and practices

## Inspect data (Slide 7)

```{r}
?msleep # metadata available on the help page (part of ggplot2 package)
glimpse(msleep) # see the data
```

## Basic plot (Slide 10)

You should understand all code used here from the previous tutorial.

```{r}
# basic plot on slide 3
p_base <- ggplot(msleep, aes(bodywt, sleep_total / awake)) +
	geom_point(aes(colour = sleep_cycle)) +
	scale_x_continuous(trans = "log") +
	scale_y_continuous(trans = "log") +
	labs(
		title = "Sleep time by body weight among mammals.",
		colour = "Sleep cycle (hr)",
		y = "Ratio of sleep time over awake time",
		x = "Body weight (kg)"
	)
p_base
```

```{r}
# practice (slide 11)
p_base +
  theme_minimal()
# add the remaining layers to the basic plot!
```

## Adjusting geometric objects (slide 13)

```{r}
p_orange <- ggplot(msleep, aes(bodywt, sleep_total / awake)) +
  geom_point(aes(colour = sleep_cycle)) +
  labs(
    title = "Sleep time by body weight among mammals.",
    colour = "Sleep cycle (hr)",
    y = "Ratio of sleep time over awake time",
    x = "Body weight (kg)"
  ) +
  scale_x_continuous(
    trans = "log", 
    breaks = c(1, 10, 100, 1000),
    labels = scales::label_number(big.mark = ",")
  ) +
  scale_y_continuous(
    trans = "log",
    breaks = c(0.1, 0.5, 1, 2.5, 5)
  ) +
  geom_smooth(
    method = "lm", 
    formula = y ~ x,
    se = FALSE, 
    linetype = "longdash", 
    colour = "darkgrey"
  ) +
  scale_colour_gradient(low = "lightgrey", high = "orange") +
  theme_minimal()
p_orange
```

```{r}
# practice (slide 14 & 15)

```

## Adding a layer with a subset (slide 17)

```{r}
# practice (slide 17)
p_missing <- ggplot(
  msleep,
  aes(bodywt, sleep_total / awake, colour = vore)
  ) +
  scale_x_continuous(
    trans = "log", 
    breaks = c(1, 10, 100, 1000),
    labels = scales::label_number(big.mark = ",")
  ) +
  scale_y_continuous(
    trans = "log",
    breaks = c(0.1, 0.5, 1, 2.5, 5)
  ) +
  labs(
    title = "Sleep time by body weight among mammals.",
    size = "Sleep cycle (hr)",
    colour = "Feeding habit",
    y = "Ratio of sleep time over awake time",
    x = "Body weight (kg)"
  ) +
   geom_smooth(
    method = "lm", 
    formula = y ~ x,
    se = FALSE, 
    linetype = "longdash", 
    colour = "darkgrey"
  ) +
  scale_colour_brewer(
    palette = "Dark2", 
    labels = c("Carnivore", "Herbivore", "Insectivore", 
               "Omnivore", "Not available"),
    na.value = "grey50"
  ) +
  theme_minimal()
```

```{r eval=FALSE}
# complete the code below to add different shaped
# points for missing data vs non-missing data
p_missing <- p_missing +
  geom_point(aes(size = sleep_cycle), alpha = 0.5,
             data = # data with sleep cycle
             ) +
  geom_point(size = 3, shape = 1, alpha = 0.5,
             # data = # data missing sleep cycle
             )
p_missing
```

## Placing legend (slide 22 & 23)

```{r}
p_full <- ggplot(msleep, aes(bodywt, sleep_total / awake, colour = vore)) +
  geom_point(aes(size = sleep_cycle), alpha = 0.5) +
  scale_x_continuous(
    trans = "log", 
    breaks = c(1, 10, 100, 1000),
    labels = scales::label_number(big.mark = ",")
  ) +
  scale_y_continuous(
    trans = "log",
    breaks = c(0.1, 0.5, 1, 2.5, 5)
  ) +
  labs(
    title = "Sleep time by body weight among mammals.",
    size = "Sleep cycle (hr)",
    colour = "Feeding habit",
    y = "Ratio of sleep time over awake time",
    x = "Body weight (kg)"
  ) +
  geom_smooth(
    method = "lm", 
    formula = y ~ x,
    se = FALSE, 
    linetype = "longdash", 
    colour = "darkgrey"
  ) +
  scale_colour_brewer(
    palette = "Dark2", 
    labels = c("Carnivore", "Herbivore", "Insectivore", 
               "Omnivore", "Not available"),
    na.value = "grey50"
  ) +
  theme_minimal()
p_full
```

```{r}
# slide 22
p_full +
  theme(
    legend.background = element_rect(fill = "white", colour = NA),
    legend.position = c(.95, .95),
    legend.justification = c(1, 1),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 6),
    legend.key.size = unit(3, "pt")
  )
```


```{r}
# slide 23
p_full +
  theme(
    legend.position = "left",
    legend.box = "horizontal",
    legend.box.just = "bottom",
    legend.margin = margin(),
    legend.justification = c(0, 0)
  )
```

```{r}
# practice ! (slide 24)
p_missing + # you must complete the code for slide 17 first!
  theme(
    # complete the code
  )
```

## Styling individual guides (slide 25)

```{r}
# you must complete the code for slide 17 first!
p_missing +
  scale_colour_brewer(
    palette = "Dark2", 
    labels = c("Carnivore", "Herbivore", "Insectivore", 
               "Omnivore", "Not available"),
    na.value = "grey50",
    guide = guide_legend(
      order = 2,
      title.position = "top",
      title.hjust = 0,
      nrow = 3,
      byrow = TRUE,
      override.aes = list(size = 3)
    )
  ) +
  scale_size_continuous(
    # practice! see slide 26
  ) +
  theme(
    legend.position = "top",
    legend.box = "horizontal",
    legend.box.just = "top",
    legend.justification = c(1, 0),
    legend.margin = margin()
  )
```

## Text styling with `ggtext` and font size (slides 28 & 35)

```{r}
# you must complete the code for slide 17 first!
p_labelled <- p_missing +
  scale_colour_brewer(
    palette = "Dark2", 
    labels = paste0(
      "<span style='color:",
      c(RColorBrewer::brewer.pal(4, "Dark2"), "#7f7f7f"), ";'>",
      c("Carnivore", "Herbivore", "Insectivore", "Omnivore", "Not available"),
      "</span>" # you can use HTML in markdown
    ),
    na.value = "grey50",
    guide = guide_legend(
      order = 2,
      title.position = "top",
      title.hjust = 0,
      byrow = TRUE,
      override.aes = list(size = 3)
    )
  ) +
  scale_size_continuous(
    breaks = c(0.5, 1, 1.5, 0.5),
    labels = c("0.5", "1", "1.5", "Not available"),
    guide = guide_legend(
      order = 1,
      title.position = "top",
      title.hjust = 0,
      label.position = "right",
      override.aes = list(shape = c(19, 19, 19, 1))
    )
  ) +
  theme(
    title = element_text(size = 24),
    legend.position = "left",
    legend.box = "vertical",
    legend.box.just = "left",
    legend.justification = c(0, 1),
    legend.margin = margin(),
    axis.title = element_text(hjust = 0, size = 16),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 12),
    # practice! slide 36
    legend.text = element_markdown() # element_markdown() available from ggtext
  )
p_labelled
```

## Storytelling & annotations (slides 39 to 41)

```{r}
ggplot(
  msleep,
  aes(bodywt, sleep_total / awake, colour = vore)
  ) +
  geom_smooth(
    method = "lm", 
    formula = y ~ x,
    se = FALSE, 
    linetype = "longdash", 
    colour = "darkgrey"
  ) +
  geom_point(aes(size = sleep_cycle), alpha = 0.5,
             data = filter(msleep, !is.na(sleep_cycle))) +
  geom_point(size = 3, shape = 1, alpha = 0.5, show.legend = FALSE,
             data = filter(msleep, is.na(sleep_cycle))) +
  scale_x_continuous(
    trans = "log", 
    breaks = c(1, 10, 100, 1000),
    labels = scales::label_number(big.mark = ",")
  ) +
  scale_y_continuous(
    trans = "log",
    breaks = c(0.1, 0.5, 1, 2.5, 5)
  ) +
  theme_minimal() +
  scale_colour_brewer(
    palette = "Dark2", 
    labels = paste0(
      "<span style='color:",
      c(RColorBrewer::brewer.pal(4, "Dark2"), "#7f7f7f"), ";'>",
      c("Carnivore", "Herbivore", "Insectivore", "Omnivore", "Not available"),
      "</span>"
    ),
    na.value = "grey50",
    guide = guide_legend(
      order = 2,
      title.position = "top",
      title.hjust = 0,
      byrow = TRUE,
      override.aes = list(size = 2)
    )
  ) +
  scale_size_continuous(
    breaks = c(0.5, 1, 1.5, 0.5),
    labels = c("0.5", "1", "1.5", "Not available"),
    guide = guide_legend(
      order = 1,
      title.position = "top",
      title.hjust = 0,
      label.position = "right",
      override.aes = list(shape = c(19, 19, 19, 1))
    )
  ) +
  theme(
    legend.text = element_markdown(size = 8),
    legend.title = element_text(size = 10),
    legend.position = "left",
    legend.box = "vertical",
    legend.box.just = "left",
    legend.justification = c(0, 1),
    legend.margin = margin(),
    legend.key.height = unit(0, "pt"),
    axis.title = element_text(hjust = 0, size = 10),
    title = element_text(size = 12),
    plot.caption = element_text(size = 8, colour = "grey50")
  ) +
  labs(
    title = "Larger mammals, specially herbivores,\ntend to sleep less, but have longer sleep cycles.",
    x = "Body weight (kg)",
    y = "Ratio of sleep time over awake time",
    size = "Sleep cycle (hr)^1",
    colour = "Feeding habit",
    caption = expression(""^1~"Sleep cycle is the period between REM seep and non-REM sleep.")
  ) +
  ## from ggrepel library
  geom_label_repel(
		label = c(
			"An average horse sleeps\nonly 2.7 hours a day!",
			"Humans sleep 8 hours a day and\nhave a sleep cycle of 1.5 hours on average.\nHope you all have 8 hours of daily sleep.", 
			"You are more likely to find\na little brown bat in their sleep\nas they sleep 19.9 hours a day."
		),
		size = 2,
		data = filter(msleep, name %in% c("Horse", "Human", "Little brown bat")),
		nudge_x = c(-1.5, -2, 1),
		nudge_y = c(-.2, -1, -.5),
		show.legend = FALSE
	)

```

## Arranging multiple plots (slides 42 & 43)

```{r}
# you can define a common theme to be used multiple times
theme_full <- theme_minimal() +
  theme(
    legend.text = element_markdown(size = 8),
    legend.title = element_text(size = 10),
    legend.position = "top",
    legend.box = "vertical",
    legend.box.just = "left",
    legend.justification = c(0, 1),
    legend.margin = margin(),
    legend.key.height = unit(0, "pt"),
    axis.title = element_text(hjust = 0, size = 10),
    title = element_text(size = 12),
    plot.caption = element_text(size = 8, colour = "grey50")
  )
# creating multipe plots to put together
full <- ggplot(mapping = aes(bodywt, sleep_total / awake)) +
  geom_smooth(
    method = "lm", se = FALSE, linetype = "longdash", colour = "darkgrey",
    data = msleep
  ) +
  geom_point(aes(size = sleep_cycle), alpha = 0.5,
             data = filter(msleep, !is.na(sleep_cycle))) +
  geom_point(size = 3, shape = 1, alpha = 0.5, show.legend = FALSE,
             data = filter(msleep, is.na(sleep_cycle))) +
  scale_x_continuous(
    trans = "log", 
    breaks = c(1, 10, 100, 1000),
    labels = scales::label_number(big.mark = ",")
  ) +
  scale_y_continuous(
    trans = "log",
    breaks = c(0.1, 0.5, 1, 2.5, 5)
  ) +
  scale_size_continuous(
    breaks = c(0.5, 1, 1.5, 0.5),
    labels = c("0.5", "1", "1.5", "Not available"),
    guide = guide_legend(
      order = 1,
      title.position = "top",
      title.hjust = 0,
      label.position = "right",
      override.aes = list(shape = c(19, 19, 19, 1))
    )
  ) +
  labs(
    title = "Larger mammals, specially herbivores, tend to sleep less, but have longer sleep cycles.",
    x = "Body weight (kg)",
    y = "Ratio of sleep time over awake time",
    size = "Sleep cycle (hr)^1",
    colour = "Feeding habit",
    caption = expression(""^1~"Sleep cycle is the period between REM seep and non-REM sleep.")
  ) +
  theme_full
pcarni <- ggplot(mapping = aes(bodywt, sleep_total / awake)) +
  geom_point(aes(size = sleep_cycle), alpha = 0.5, show.legend = FALSE,
             data = filter(msleep, !is.na(sleep_cycle), vore == "carni")) +
  geom_point(size = 3, shape = 1, alpha = 0.5, show.legend = FALSE,
             data = filter(msleep, is.na(sleep_cycle), vore == "carni")) +
  scale_x_continuous(
    trans = "log", 
    breaks = c(1, 10, 100, 1000),
    labels = scales::label_number(big.mark = ",")
  ) +
  scale_y_continuous(
    trans = "log",
    breaks = c(0.1, 0.5, 1, 2.5, 5)
  ) +
  scale_size_continuous(
    breaks = c(0.5, 1, 1.5, 0.5),
    labels = c("0.5", "1", "1.5", "Not available"),
    guide = guide_legend(
      order = 1,
      title.position = "top",
      override.aes = list(shape = c(19, 19, 19, 1))
    )
  ) +
  theme_full +
  labs(subtitle = "Carnivores", x = "", y = "")
pomni <- ggplot(mapping = aes(bodywt, sleep_total / awake)) +
  geom_point(aes(size = sleep_cycle), alpha = 0.5, show.legend = FALSE,
             data = filter(msleep, !is.na(sleep_cycle), vore == "omni")) +
  geom_point(size = 3, shape = 1, alpha = 0.5, show.legend = FALSE,
             data = filter(msleep, is.na(sleep_cycle), vore == "omni")) +
  scale_x_continuous(
    trans = "log", 
    breaks = c(1, 10, 100, 1000),
    labels = scales::label_number(big.mark = ",")
  ) +
  scale_y_continuous(
    trans = "log",
    breaks = c(0.1, 0.5, 1, 2.5, 5)
  ) +
  scale_size_continuous(
    breaks = c(0.5, 1, 1.5, 0.5),
    labels = c("0.5", "1", "1.5", "Not available"),
    guide = guide_legend(
      order = 1,
      title.position = "top",
      title.hjust = 0,
      label.position = "right",
      override.aes = list(shape = c(19, 19, 19, 1))
    )
  ) +
  theme_full +
  labs(subtitle = "Omnivores", x = "", y = "")
pinsecti <- ggplot(mapping = aes(bodywt, sleep_total / awake)) +
  geom_point(aes(size = sleep_cycle), alpha = 0.5, show.legend = FALSE,
             data = filter(msleep, !is.na(sleep_cycle), vore == "insecti")) +
  geom_point(size = 3, shape = 1, alpha = 0.5, show.legend = FALSE,
             data = filter(msleep, is.na(sleep_cycle), vore == "insecti")) +
  scale_x_continuous(
    trans = "log", 
    breaks = c(1, 10, 100, 1000),
    labels = scales::label_number(big.mark = ",")
  ) +
  scale_y_continuous(
    trans = "log",
    breaks = c(0.1, 0.5, 1, 2.5, 5)
  ) +
  scale_size_continuous(
    breaks = c(0.5, 1, 1.5, 0.5),
    labels = c("0.5", "1", "1.5", "Not available"),
    guide = guide_legend(
      order = 1,
      title.position = "top",
      title.hjust = 0,
      label.position = "right",
      override.aes = list(shape = c(19, 19, 19, 1))
    )
  ) +
  theme_full +
  labs(subtitle = "Insectivores", x = "", y = "")
pherbi <- ggplot(mapping = aes(bodywt, sleep_total / awake)) +
  geom_point(aes(size = sleep_cycle), alpha = 0.5, show.legend = FALSE,
             data = filter(msleep, !is.na(sleep_cycle), vore == "herbi")) +
  geom_point(size = 3, shape = 1, alpha = 0.5, show.legend = FALSE,
             data = filter(msleep, is.na(sleep_cycle), vore == "herbi")) +
  scale_x_continuous(
    trans = "log", 
    breaks = c(1, 10, 100, 1000),
    labels = scales::label_number(big.mark = ",")
  ) +
  scale_y_continuous(
    trans = "log",
    breaks = c(0.1, 0.5, 1, 2.5, 5)
  ) +
  scale_size_continuous(
    breaks = c(0.5, 1, 1.5, 0.5),
    labels = c("0.5", "1", "1.5", "Not available"),
    guide = guide_legend(
      order = 1,
      title.position = "top",
      title.hjust = 0,
      label.position = "right",
      override.aes = list(shape = c(19, 19, 19, 1))
    )
  ) +
  theme_full +
  labs(subtitle = "Herbivores", x = "", y = "")

# patchwork library allows arranging multiple plots
patched <- full + 
  (pcarni + pomni + pinsecti + pherbi + plot_layout(nrow = 1)) +
  plot_layout(ncol = 1)

patched
```

# Exercises

```{r}
# start by getting the subset of data you wish to focus on
human <- starwars |>
  filter(species == "Human") |>
  select(name, skin_color, hair_color, gender)
# films and starships are nested columns and requires "unnest"ing
film_and_ships <- starwars |>
  select(name, films, starships) |>
  unnest(films, keep_empty = TRUE) |>
  unnest(starships, keep_empty = TRUE)
film_year <- data.frame(
  films = film_and_ships |> distinct(films),
  year = c(1980, 2005, 1983, 1977, 2015, 2002, 1999)
)
starwars_ex <- human |>
  left_join(film_and_ships, by = "name") |>
  left_join(film_year, by = "films") |>
  mutate(
    skin_color = fct_relevel(
      skin_color, c("pale", "light", "white", "fair", "tan", "dark"))
  )
glimpse(starwars_ex)
```

## Exercise 1

> Suppose you want to display how the proportion of feminine human characters have changed over the years. Decide on how you would summarize the data and create a plot with appropriate design. 

```{r}

```

## Exercise 2

> You notice feminine human characters are rarely on a starship in the earlier movies and you want to display gender, movie, and whether they were on a starship on a single plot.
Construct a plot with appropriate legends. You can refer to the plot on the right or come up with your own design.

```{r}

```

## Exericse 3

> Arrange the plots from Exercise 1, Exercise 2, and/or any other plots together to come up with a coherent insight about how gender diversity has evolved (or not evolved) in Starwars movies.
Use appropriate title, annotations, etc. as desired.

```{r}

# save the resulting plot; adjust the dimensions as needed
# ggsave("starwars_ex.png") 
```
