---
title: "Shiny & ggiraph tutorial"
date: "2024-10-11"
output: 
  html_document:
    toc: true
    theme: united
runtime: shiny
---

# Shiny & ggiraph Tutorial

## Setup and Data

```{css, echo=FALSE}
.shiny-frame{height: 680px;} # set shiny app height
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
required_packages <- c("shiny", "ggiraph", "ggplot2",
                       "patchwork", "sf", "tidyverse", "viridis")
for (p in required_packages) {
  if (!require(p, character.only = TRUE)) {
    install.packages(p, character.only = TRUE)
  }
}
library(tidyverse)
library(shiny)
library(ggiraph)
library(ggplot2)
library(patchwork)
library(sf)
```

```{r Data}
head(msleep)
glimpse(msleep)
help(msleep)
```

```{r Preprocessing}
data("msleep", package = "ggplot2")

# Add binary sleep cycle variable to the dataset
msleep <- msleep |>
  mutate(
    sleep_cycle_bi = if_else(
      sleep_cycle > median(sleep_cycle, na.rm = TRUE),
      "Above Median", 
      "Below Median"
    )
  )
```

## Static ggplot graph

```{r Static, echo=FALSE}
static_plot <- msleep |> 
  mutate(
    # create the binary sleep cycle variable
    sleep_cycle_bi = sleep_cycle > median(sleep_cycle, na.rm = TRUE) 
  ) |>
  ggplot(aes(bodywt, sleep_total / awake)) +
  # use the binary variable to colour points only
  geom_point(aes(colour = sleep_cycle_bi)) + 
  scale_x_continuous( # set limits and breaks on log-transformed axes
    trans = "log", 
    breaks = c(1, 10, 100, 1000),
    labels = scales::label_number(big.mark = ",")
  ) +
  scale_y_continuous( # set limits and breaks on log-transformed axes
    trans = "log",
    breaks = c(0.1, 0.5, 1, 2.5, 5)
  ) +
  labs(
    title = "Sleep time by body weight among mammals.",
    colour = "Sleep cycle",
    y = "Ratio of sleep time over awake time",
    x = "Body weight (kg)"
  ) +
  geom_smooth( # add regression line
    formula = "y ~ x",
    method = "lm", 
    se = FALSE, # linear line without the confidence band
    linetype = "longdash", 
    colour = "darkgrey"
  ) +
  scale_colour_viridis_d(
    begin = .1,
    end = .7,
    na.value = "grey80",
    labels = c("Above Median", "Below Median", "Not Available"),
    guide = guide_legend(override.aes = list(fill = c("white", "white", "grey80")))
  ) +
  theme_minimal()

static_plot
```

## Shiny and ggiraph

In this example we go over three main ideas: 

1. Selection Interactions (hover, selection, and Shiny dropdown):
  + `girafeOutput(<plot-name>)`: Creates an output placeholder for displaying the interactive plot.
  + `selectInput(inputId = <input-id>, choices, ...)`: Adds a dropdown menu to filter the data by cylinder, we also need to add corresponding logic to filter the data based on the selection
  + `geom_point_interactive(...)`: Adds points with interactivity enabled for hover and selection effects.
  + `opts_hover/opts_selection(...)`: Sets up a hover/selection effect to highlight the slected data point.
  
2. Detail Interactions (tooltips on hover and print selected points):
  + `tooltip = paste(<text-you-want-to-show>)`: Creates a tooltip.
  + `textOutput(<text-output-name>)`: Adds a text output area in the UI.
  + `output$<text-output-name> <- renderPrint({...})`: Prints out the details of the selected point when clicked.
  + `input$<plot-name>_selected`: Get the data_id for selected element

3. Navigation Interactions(Zoom):
  + `opts_zoom(min = 1, max = 2)`: Sets up zoom interaction with a minimum zoom level of 1x and a maximum of 2x.

```{r Shiny and ggiraph}
# Define the UI for the Shiny app
ui <- fluidPage( # Fluid layout for responsive design
  titlePanel("Interactive Sleep Time vs Body Weight Plot"),
  
  sidebarLayout( # Sidebar layout with a sidebar panel and main panel
    sidebarPanel( # Sidebar Panel
      # Create a dropdown for selecting the sleep cycle group (optional filtering)
      selectInput(
        inputId = "cycle_filter",
        label = "Filter by Sleep Cycle:",
        choices = c("All", "Below Median", "Above Median"),
        selected = "All"
      ),
      # Show the name of the clicked point
      textOutput("clicked_point")
    ),
    mainPanel( # Main Panel
      # Use girafeOutput to display the interactive plot
      girafeOutput("interactive_plot", width = "100%", height = "600px")
    )
  )
)

# Define the server logic for the Shiny app
server <- function(input, output) {
  
  # Create the interactive plot using ggiraph
  output$interactive_plot <- renderGirafe({
    
    # Build the ggplot with interactive elements
    chart_interactive <- msleep |> 
      ggplot(
        aes(
          x = bodywt, 
          y = sleep_total / awake, 
          colour = sleep_cycle_bi,
          fill = sleep_cycle_bi,
          size = sleep_cycle_bi,
          ##### required for interaction #####
          # Set the data_id to the 'name' column to uniquely identify each point
          data_id = name,
          tooltip = paste(
            "Name:", name, 
            "<br />Body Weight:", bodywt, "kg",
            if_else(
              is.na(sleep_cycle_bi),
              "",
              paste0(
                "<br />Sleep Cycle: ",
                "<b style='",
                "text-shadow: .2px .2px 0 white, -.2px -.2px 0 white, ",
                ".2px -.2px 0 white, -.2px .2px 0 white;",
                "color:", 
                # get matching colours from viridis::viridis(2, begin = .1, end = .7)
                if_else(sleep_cycle_bi == "Above Median", "#CF9FFF", "#43BF71FF"),
                "';>", sleep_cycle_bi, "</b>"
              )
            )
          )
        )
      ) +
      # Adds interactive points for hover and selection effects
      # TODO: Adds interactive points for hover and selection effects, set shape = 21
       +
      scale_x_continuous(
        trans = "log", 
        breaks = c(1, 10, 100, 1000),
        labels = scales::label_number(big.mark = ",")
      ) +
      scale_y_continuous(
        trans = "log",
        breaks = c(0.1, 0.5, 1, 2.5, 5)
      ) +
      scale_fill_manual(
        values = alpha(viridis::viridis(2, begin = .1, end = .7), .3),
        na.value = alpha("grey90", .3),
        guide = NULL
      ) +
      labs(
        title = "Sleep Time by Body Weight among Mammals",
        colour = "Sleep Cycle",
        y = "Ratio of Sleep Time over Awake Time",
        x = "Body Weight (kg)"
      ) +
      geom_smooth(
        formula = y ~ x,
        method = "lm", 
        se = FALSE,
        linetype = "longdash", 
        colour = "darkgrey"
      ) +
      # viridis colour palette for discrete variables
      scale_colour_viridis_d_interactive(
        tooltip = function(x) if_else(is.na(x), "Missing sleep cycle data", x),
        begin = .1,
        end = .7,
        na.value = "grey80",
        labels = c("Above Median", "Below Median", "Not Available"),
        guide = guide_legend(override.aes = list(fill = c("white", "white", "grey80")))
      ) +
      theme_minimal()
    
    
    # Filter data based on the sleep cycle selection from the dropdown
    if (input$cycle_filter == "All") {
      chart_interactive <- chart_interactive +
        scale_size_manual(
          values = c(4, 4),
          na.value = 4,
          guide = NULL
        )
    } else {
      chart_interactive <- chart_interactive +
        scale_size_manual(
          values = if_else(
            c("Above Median", "Below Median", "") == input$cycle_filter, NA, 4
          ),
          guide = NULL
        )
    }
    
    # Render the plot as a girafe object
    girafe(
      ggobj = chart_interactive,
      options = list(
        # TODO: Allow single selection
        # Hint: check the documentation for opts_selection
        opts_selection(css = "stroke: black; stroke-width: 1px; fill: green;"),
        # TODO: Set up hover options, feel free to implement any css style
        opts_hover(), 
        # TODO: Set up zoom options
        opts_zoom(),
        # semi-transparent rounded tooltip
        opts_tooltip(
          css = "border-radius: 5px; background-color: rgba(0,0,0,0.7); color: white; padding: 5px;"
        )
      )
    )
  })
  
  # Show the name of the clicked point using a text output
  #TODO: Please fill the missing blank after output$
  output$ <- renderText({
    if (is.null(input$interactive_plot_selected)) {
      return("Click on a point to see its name.")
    } else {
      #TODO: Please fill the missing blank after input$
      paste("You selected:", input$)
    }
  })
}

# Run the Shiny app
shinyApp(ui = ui, server = server)
```

## Connected graphs with patchwork

New additions:

1. `combined_plot <- (<plot 1> + <plot 2>) /<plot 3>`: Combines the scatter plot (p1), bar chart (p2), and choropleth map (p3) into a single patchwork layout.
2. `geom_sf_interactive(data=<geom-data>, ...)`: Creates interactive spatial data by integrating geom_sf with interactive features.

```{r Patchwork}
# Read the full world map
world_sf <- read_sf("https://raw.githubusercontent.com/holtzy/R-graph-gallery/master/DATA/world.geojson")
world_sf <- world_sf |>
  filter(!name %in% c("Antarctica", "Greenland"))

# Create a sample dataset
set.seed(123)
happiness_data <- data.frame(
  Country = c(
    "France", "Germany", "United Kingdom",
    "Japan", "China", "Vietnam",
    "United States of America", "Canada", "Mexico"
  ),
  Continent = c(
    "Europe", "Europe", "Europe",
    "Asia", "Asia", "Asia",
    "North America", "North America", "North America"
  ),
  Happiness_Score = rnorm(mean = 30, sd = 20, n = 9),
  GDP_per_capita = rnorm(mean = 30, sd = 20, n = 9),
  Social_support = rnorm(mean = 30, sd = 20, n = 9),
  Healthy_life_expectancy = rnorm(mean = 30, sd = 20, n = 9)
)

# Join the happiness data with the full world map
world_sf <- world_sf |>
  left_join(happiness_data, by = c("name" = "Country"))

# Define the Shiny UI
ui <- fluidPage(
  titlePanel(
    div(
      p("World's Happiness Data Visualization"),
      em("(Numbers are randomly generated)")
    )
  ),
  
  mainPanel(
    # Use girafeOutput to display the interactive plot
    girafeOutput("patchwork_plot", width = "100%", height = "600px"),
  )
)

server <- function(input, output) {
  
  # Create the interactive plot using ggiraph
  output$patchwork_plot <- renderGirafe({
    
     world_sf_happy <- world_sf |>
       filter(!is.na(Happiness_Score))

    # Create the first chart (Scatter plot)
    p1 <- world_sf_happy |>
      ggplot() +
      geom_point_interactive(
        aes(
          GDP_per_capita,
          Happiness_Score,
          tooltip = name,
          data_id = name,
          color = name
        ), 
        size = 4,
        alpha = .7
      ) +
      scale_color_brewer(palette = "Set1") +
      theme_minimal() +
      theme(
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "none"
      )
    
    # Create the second chart (Bar plot)
    p2 <- world_sf_happy |>
      ggplot(
        aes(
          x = Happiness_Score,
          y = reorder(name, Happiness_Score),
          tooltip = name,
          data_id = name,
          fill = name
        )
      ) +
      geom_col_interactive(alpha = .7, orientation = "y") +
      scale_fill_brewer(palette = "Set1") +
      theme_minimal() +
      theme(
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "none"
      )
    
    # Create the third chart (choropleth)
    p3 <- ggplot() +
      # draw the borders of the "whole" world
      geom_sf(data = world_sf, fill = "white", color = "grey50", linewidth = .5) +
      geom_sf_interactive(
        data = world_sf_happy,
        aes(fill = name, tooltip = name, data_id = name),
        alpha = .7
      ) +
      # coordinate reference system (commonly used is 3857 - e.g., Google Maps)
      coord_sf(crs = st_crs(4326)) + # used for Google Earth
      theme_void() +
      scale_fill_brewer(palette = "Set1") +
      theme(
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "none"
      )
    
    # Combine the plots
    combined_plot <- (p1 + p2) / p3 + plot_layout(heights = c(1, 2))
    # Plus means horizontal alignment
    # Forward Slash is Vertical alignment
    
    # Create the interactive plot
    patchwork_plot <- girafe(ggobj = combined_plot)
    patchwork_plot <- girafe_options(
      patchwork_plot,
      opts_selection(css="fill: #2AFEB7"),
      opts_hover(css = "fill-opacity:1; stroke: black;"),
      opts_zoom(min = 1, max = 2)
    )
  })
}

shinyApp(ui = ui, server = server)
```

## Exercise Template

```{r Data Preparation}
# loading data
casualties <- read_csv("data/nigthingale.csv") |>
  mutate(
    # converts the Date string into a "date" object 
    # with which you can use ?scale_*_date() in ggplot
    date = as_date(Date, format = "%m/%Y")
  ) 
# create your visualization here
# you can save your plot by calling ggplot() after generatin each plot
casualties <- casualties %>% 
  mutate(Year = format(date, "%Y"))

casualties
```

```{r Excercise: Static Plot}
casualties |>
  mutate(rate = Wounds / Army * 1000) |>
  ggplot(aes(x = date, y = rate)) +
  theme_classic() +
  geom_line() +
  geom_point(aes(colour = rate == max(rate), size = rate == max(rate))) +
  scale_colour_manual(values = c(`TRUE` = 2, `FALSE` = "grey50"),
                      guide = NULL) +
  scale_size_manual(values = c(`TRUE` = 5, `FALSE` = 2),
                    guide = NULL) +
  scale_y_continuous(breaks = c(0, 5, 10)) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b, %Y") +
  labs(
    x = NULL, 
    # \n creates a line break
    y = "Casualty rate in the field\n(per 1,000)"
  ) 
```

Convert the static plot we used for week 2's tutorial to interactive plot by using the interactive components introduced in today's tutorial(Click, Hover, Dropdown, Tooltips, Zoom). Be creative!

```{r Exercise: Interative Plot}
ui <- fluidPage( # Fluid layout for responsive design
  
)

# Define the server logic for the Shiny app
server <- function(input, output) {
  
}

# Run the Shiny app
shinyApp(ui = ui, server = server)
```

## Resources

### Shiny (for R)

- [Getting Started](https://shiny.posit.co/r/getstarted/shiny-basics/lesson1/index.html)
- [Gallery](https://shiny.rstudio.com/gallery/)
- [Cheatsheet](https://shiny.posit.co/r/articles/start/cheatsheet/)
- [Mastering Shiny Book](https://mastering-shiny.org/)
- [Comprehensive Guides](https://shiny.rstudio.com/articles/)

### ggiraph

- [Overview](https://davidgohel.github.io/ggiraph/)
- [ggiraph Book](https://www.ardata.fr/ggiraph-book/)
- [Stackoverflow](https://stackoverflow.com/questions/tagged/ggiraph?tab=Votes)
