---
title: "A tour of ggplot"
author: "STA313"
date: "2024-09-13"
output: html_document
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require(palmerpenguins)) {
  install.packages("palmerpenguins") # install the package if missing
}
if (!require(mosaicData)) {
  install.packages("mosaicData") # install the package if missing
}
library(tidyverse) 
library(mosaicData)
data(SAT)
```

I am loading all required libraries in the `setup` code chunk above to 

1. ensure all required libraries are available throughout the document; and
2. above loading the packages multiple times unnecessarily.

Alternatively, you can load a library in a separate code chunk just before you start
using it. 

_Note: Loading `tidyverse` also loads `ggplot2`. We will also use functions
from the `tidyverse` family which are not included in `ggplot2`. Alternatively, 
you can individually load `ggplot2` and `dplyr` which contains the other required functions._


## Palmer penguins

Measurements for penguins species, island in Palmer Archipelago, size (flipper length, body mass, bill dimensions), and sex.

```{r}
glimpse(penguins)
```

```{r warning=FALSE}
# bill depth and length plot
ggplot(
  penguins,
  aes(x = bill_depth_mm, y = bill_length_mm, colour = species)
) +
  geom_point() +
  labs(
    title = "Bill sizes may help distinguish different penguin species",
    subtitle = "Bill depths and lengths of Adelie, Chinstrap, and Gentoo penguins",
    x = "Bill depth (mm)",
    y = "Bill length (mm)",
    colour = "Species"
  )
```

## Application exercise 1

> Using the `mpg` data set, map the engine displacement values to the x-ais and the highway miles per gallon to the y-axis. Represent each observation with a point. Apply the "darkgreen" colour to the entire geometry. Add a meaningful title (and subtitle), and labels.

First, have a look at the data set. Alternatively, type `?mpg` in the console to 
load the documentation page for more details after loading `tidyverse`.

```{r}
glimpse(mpg) # mpg and glimpse are available through `tidyverse`
```


```{r}
ggplot(mpg, aes(x = displ, y = hwy))
# complete the plot
```


## Application exercise 2

> Using the `mpg` data set, recreate the chart below. (See slide 39)

```{r}
ggplot(mpg) + # complete the mapping
  geom_bar(fill = "orange")         # geom_bar is used to plot bars
# complete the plot
```

## Application exercise 3

> Using the `mpg` data set, recreate the chart below. (See slide 55)

```{r}
mpg |>
  mutate(class = forcats::fct_infreq(class)) |> 
  ggplot() +
  geom_bar(aes(class), fill = "orange")
```

_Note: The pipe operator `|>` feeds the output of the left-hand side to the call on the right-hand side as its first argument. In the code chunk above,  the output of `mutate(mpg, class = forcats::fct_infreq(class))` is used as the first (`data`) argument of `ggplot()` call. `|>` is available from R version 4.0+._

## Application exercise 4

```{r eval=FALSE}
# included for demonstration purposes but not executed
# the data is loaded at the top of the document already
library(mosaicData)
data(SAT)
```

> Reproduce this plot following the steps on the next slide. 
(See slides 65 and 66).

First, have a look at the data set. Alternatively, type `?SAT` in the console to 
load the documentation page for more details after loading `mosaicData`.

```{r}
glimpse(SAT) # SAT is available through `mosaicData`
```

```{r}
# create the plot
```

## Application exercise 5

> Using the `mpg` data set, recreate the chart below. (See slide 67)

```{r}
mutate(mpg, class = forcats::fct_infreq(class)) |>
  # complete the plot
  scale_fill_viridis_d()           # see ?scale_fill_viridis_d
```


_Note: If you want to sort the values based on other criteria, see `?forcats::fct_infreq` for other related functions. Alternatively, you can use `levels` argument in `factor()` function to manually order the values of a categorical variable._
